---
title: "RepData"
author: "Mark Kaghazgarian"
date: "11/21/2017"
output: 
  html_document: 
    fig_caption: yes
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(fig.keep = "all", fig.path = "figure/") 
```

## Loading and preprocessing the data

```{r loadData, echo=TRUE}
library("data.table")
library("ggplot2")
df <- read.csv(unz("./activity.zip", "activity.csv"))
dt <- data.table(df)
```

## What is mean total number of steps taken per day?
```{r mean, echo=TRUE}
dt[, mean(steps, na.rm = TRUE), by = date]
```

```{r hist1, echo=TRUE}
dt_hist <- dt[, .(steps = sum(steps, na.rm = TRUE)), by = date]
hist(dt_hist$steps, 
     breaks = 8, 
     col = "blue", 
     main="Frequnecy of total steps in each day", 
     xlab = "Steps")
```

```{r}
dt_hist[, mean(steps)]
dt_hist[, median(steps)]
```

## What is the average daily activity pattern?

First evaluate the average number of steps taken in each interval:

```{r}
steps_avg <- lapply(with(dt, split(steps,interval)),mean, na.rm = TRUE)
```

Then plot based on result of above calculation:
```{r plot1, echo=TRUE}
plot(names(steps_avg), steps_avg, type = "l",
     main= "Average taken steps over all days",
     xlab="Interval",
     ylab="Average steps")
```

In order to find interval which, on average across all the days in the dataset, contains the maximum number of steps:

```{r max, echo=TRUE}
steps_avg[which.max(steps_avg)]
```

## Imputing missing values

 * For counting **NA** in dataset we can do like this:
 
```{r}
apply(dt, 2, function(x) sum(length(which(is.na(x)))))
```

 * In order to impute the missing values in steps, I decided to use the mean of taken steps in the same interval in all days:
 
```{r}
## First we need to cast steps to double in order to prevent auto 
## casting from double to int

dt[, steps := as.double(steps)]

## Then, we need to iterate over the dataset and fill the missing
## values with mean values
for (i in 1:nrow(dt)) {
    if (is.na(dt[i,steps])) {
        dt[i,steps := as.double(steps_avg[[as.character(dt[i,interval])]])]
    }
}
```

Plot a histogram of new constructed dataset imputed with mean values:

```{r hist2, echo=TRUE}
dt_hist <- dt[, .(steps = sum(steps, na.rm = TRUE)), by = date]
hist(dt_hist$steps, 
     breaks = 8, 
     col = "blue", 
     main="Frequnecy of total steps in each day (Imputed dataset version)", 
     xlab = "Steps")
```

**Mean** and **Median** total number of steps taken per day after filling missed value with means:
```{r}
dt_hist[, mean(steps)]
dt_hist[, median(steps)]
```
 
## Are there differences in activity patterns between weekdays and weekends? 
 
```{r}
dt[, weekend := ifelse (weekdays(as.Date(date)) %in% c("Saturday", "Sunday") , "weekend" , "weekday")]
```

```{r}
dt2 <- dt[,.(steps = mean(steps),weekend,interval), by=c('interval','weekend')]

```

```{r plot2, echo=TRUE}
qplot(interval, steps, 
     data = dt2,
     geom=c("line"),
     xlab = "Interval", 
     ylab = "Number of steps", 
     main = "") + facet_wrap( ~ weekend, ncol = 1)
```

